---
format: gfm
editor:
  markdown:
    wrap: 80
---

# Functions (part 2)

## Euclidean distance between two points

This function returns the distance from a reference point in the real line
to 1,2, ..., n points in the real line.

```cpp
[[cpp11::register]] doubles pdist_cpp_(double x, doubles ys) {
  int n = ys.size();
  writable::doubles out(n);
  for(int i = 0; i < n; ++i) {
    out[i] = sqrt(pow(ys[i] - x, 2.0));
  }
  return out;
}
```

The R equivalent is the following.

```r
#' Return the euclidean distance from a value to a vector of values (R)
#' @param x numeric
#' @param y numeric vector
#' @export
pdist_r <- function(x, ys) {
  sqrt((x - ys) ^ 2)
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the euclidean distance from a value to a vector of values (C++)
#' @inheritParams pdist_r
#' @export
pdist_cpp <- function(x, ys) {
  pdist_cpp_(x, ys)
}
```

A benchmark of the two functions is the following.

```{r 03-functions-bench}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

x <- runif(1e3) # 1,000,000 elements

# print the first 10 output's coordinates
head(pdist_cpp(0.5, x))
head(pdist_r(0.5, x))

mark(
    pdist_cpp(0.5, x),
    pdist_r(0.5, x)
)
```

Once again, R wins (for now).

## Average the elements of a vector

This function returns the average (or mean) of the elements of a vector.

```cpp
[[cpp11::register]] double mean_cpp_(doubles x) {
  int n = x.size();
  double y = 0;

  for(int i = 0; i < n; ++i) {
    // y += x[i] / n;
    // replace N divisions with 1 division at the end
    y += x[i];
  }
  
  // return y;
    return y / n;
}
```

The R equivalent is the following.

```r
#' Return the mean of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
mean_r <- function(x) {
  sum_r(x) / length(x)
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the mean of the coordinates of a vector (C++)
#' @inheritParams mean_r
#' @export
mean_cpp <- function(x) {
  mean_cpp_(x)
}
```

A benchmark of the two functions is the following.

```{r 03-functions-bench2}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

x <- runif(1e3) # 1,000,000 elements

mean(x)
mean_cpp(x)
mean_r(x)

mark(
    mean(x),
    mean_cpp(x),
    mean_r(x)
)
```

I got ties. My C++ function is clearly faster than my R function, but is
marginally slower than R's `mean()` function.

## Cumulative sum of the elements of a vector

To be able to create writable vectors, I need to append this below the namespace
declaration (i.e., `using namespace cpp11`).

```cpp
namespace writable = cpp11::writable;
```

This function returns the cumulative sum of the elements of a vector.

```cpp
[[cpp11::register]] doubles cumsum_cpp_(doubles x) {
  int n = x.size();
  writable::doubles out(n);

  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] + x[i];
  }
  return out;
}
```

The R equivalent is the following.

```r
#' Return the cumulative sum of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
cumsum_r <- function(x) {
  n <- length(x)
  out <- numeric(n)
  out[1] <- x[1]
  for (i in 2:n) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the cumulative sum of the coordinates of a vector (C++)
#' @inheritParams cumsum_r
#' @export
cumsum_cpp <- function(x) {
  cumsum_cpp_(as.double(x))
}
```

A benchmark of the two functions is the following.

```{r 03-functions-bench3}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

x <- runif(1e3) # 1,000,000 elements

cumsum(1:3)
cumsum_cpp(1:3)
cumsum_r(1:3)

mark(
    cumsum(x),
    cumsum_cpp(x),
    cumsum_r(x)
)
```

My C++ function is in the middle between R's `cumsum()` and my R function
`cumsum_r()`.

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
