# Statistical functions

```{r 03-setup}
#| label: setup
#| echo: false
#| message: false
load_all()
```

## Fair warning

These functions ignore `NA` values. These are altered in the 6th chapter of
these notes.

## Sum of Vector Elements (`sum()`)

I know R provides the `sum()` function, but I want to create this function
just for the sake of the example.

The following function returns the sum of the elements of a vector.

```cpp
[[cpp11::register]] double sum_cpp_(doubles x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}
```

The R equivalent is the following.

```r
#' Return the sum of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
sum_r <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
```

I also have to add the corresponding auxiliary function for the documentation.

```r
#' Return the sum of the coordinates of a vector (C++)
#' @inheritParams sum_r
#' @export
sum_cpp <- function(x) {
  sum_cpp_(x)
}
```

To test, I run the following lines in the R console.

```{r 03-sum-test}
set.seed(123) # for reproducibility
x <- runif(1e3) # 1,000,000 elements

sum(x)
sum_cpp(x)
sum_r(x)

mark(
  sum(x),
  sum_cpp(x),
  sum_r(x)
)
```

This is not computationally efficient (i.e., see the median time, which is
median execution time for repeated runs), which is fine because the purpose of
this example is to show how to create a package with C++ code. The code
optimizations were deactived in the `Makevars` file and the improvements to the
code will come later.

## Arithmetic Mean (`mean()`)

This function returns the average (or mean) of the elements of a vector.

```cpp
[[cpp11::register]] double mean_cpp_(doubles x) {
  int n = x.size();
  double y = 0;

  for(int i = 0; i < n; ++i) {
    // y += x[i] / n;
    // replace N divisions with 1 division at the end
    y += x[i];
  }
  
  // return y;
    return y / n;
}
```

The R equivalent is the following.

```r
#' Return the mean of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
mean_r <- function(x) {
  sum_r(x) / length(x)
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the mean of the coordinates of a vector (C++)
#' @inheritParams mean_r
#' @export
mean_cpp <- function(x) {
  mean_cpp_(x)
}
```

A benchmark of the two functions is the following.

```{r 03-mean-bench}
mean(x)
mean_cpp(x)
mean_r(x)

mark(
  mean(x),
  mean_cpp(x),
  mean_r(x)
)
```

My C++ function is faster than my R function, but is slower than R's 
`mean()` function.

## Variance (`var()`)

This function returns the variance of the elements of a vector.

```cpp
[[cpp11::register]] double var_cpp_(doubles x) {
  int n = x.size();

  if (n < 2) {
    return NA_REAL;
  }

  double ex = 0;
  for (int i = 0; i < n; ++i) {
    ex += x[i];
  }
  ex /= n;

  double out = 0;
  for (int i = 0; i < n; ++i) {
    out += pow(x[i] - ex, 2);
  }

  return out / (n - 1);
}
```

The R equivalent is the following.

```r
#' Return the variance of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
var_r <- function(x) {
  n <- length(x)
  x1 <- 0
  x2 <- 0
  for (i in seq_along(x)) {
    x1 <- x1 + x[i]
    x2 <- x2 + x[i]^2
  }
  (x2 - x1^2 / n) / (n - 1)
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the variance of the coordinates of a vector (C++)
#' @inheritParams var_r
#' @export
var_cpp <- function(x) {
  var_cpp_(x)
}
```

A benchmark of the two functions is the following.

```{r 03-var-bench}
var(x)
var_cpp(x)
var_r(x)

mark(
  var(x),
  var_cpp(x),
  var_r(x)
)
```

My C++ function is faster than my R function, but is slower than R's 
`var()` function.

## Root Mean Square Error

The next function returns the measure of the differences between the observed
values or an estimator (x1, x2, ..., xn) and the true value (x0). For example,
x1, ..., xn could be experimental averages and x0 the true average.

```cpp
[[cpp11::register]] double rmse_cpp_(doubles x, double x0) {
  int n = x.size();
  double out;
  for(int i = 0; i < n; ++i) {
    out += pow(x[i] - x0, 2.0);
  }
  return sqrt(out / n);
}
```

The R equivalent is the following.

```r
#' Return the root mean square error (R)
#' @param x numeric vector
#' @param x0 numeric value
#' @export
rmse_r <- function(x, x0) {
  sqrt(sum((x - x0)^2) / length(x))
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the root mean square error (C++)
#' @inheritParams rmse_r
#' @export
rmse_cpp <- function(x, x0) {
  rmse_cpp_(x, x0)
}
```

A benchmark of the two functions is the following.

```{r 03-functions-bench}
# create a list with 100 normal distributions with mean 0 and 1 million elements
# each
set.seed(123) # for reproducibility
x <- list()
for (i in 1:1e3) {
  x[[i]] <- rnorm(1e3)
}

# compute the mean of each distribution
x <- sapply(x, mean)

rmse_cpp(x, 0)
rmse_r(x, 0)

mark(
  rmse_cpp(x, 0),
  rmse_r(x, 0)
)
```

The R implementation is faster than the C++ implementation.

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
