---
format: gfm
editor:
  markdown:
    wrap: 80
---

# Functions (part 3)

## Logical tests

The following function is the C++ implementation of the R function `any`,
which returns `TRUE` if there is any `TRUE` element in a vector, and `FALSE`
otherwise.

```cpp
[[cpp11::register]] bool any_cpp_(logicals x) {
  int n = x.size();

  for(int i = 0; i < n; ++i) {
    if (x[i]) {
      return true;
    }
  }
  return false;
}
```

The R equivalent is the following.

```r
#' Return TRUE if any element in a vector is TRUE (R)
#' @param x logical vector
#' @export
any_r <- function(x) {
  n <- length(x)

  for (i in 1:n) {
    if (x[i]) {
      return(TRUE)
    }
  }
  FALSE
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return TRUE if any element in a vector is TRUE (C++)
#' @inheritParams all_r
#' @export
any_cpp <- function(x) {
  any_cpp_(x)
}
```

A benchmark of the two functions is the following.

```{r 04-functions-bench2}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements
y <- ifelse(x > 2, TRUE, FALSE)

any(y)
any_cpp(y)
any_r(y)

mark(
  any(y),
  any_cpp(y),
  any_r(y)
)
```

## Logical tests (cont'd)

The next function is an alternative C++ implementation of the R function
`which`, that returns the indexes of the `TRUE` elements in a vector.

```cpp
[[cpp11::register]] integers which_cpp_(cpp11::function pred, doubles x) {
  int n = x.size();
  writable::doubles res;
  int j = 0;

  for(int i = 0; i < n; ++i) {
    if (pred(x[i])) {
      ++j;
      res.push_back(i + 1);
    }
  }

  if (j == 0) {
    return doubles(0);
  } else {
    return res;
  }
}
```

The R equivalent is the following.

```r
#' Return the index of the first TRUE element in a vector (R)
#' @param x predicate function
#' @param y vector of values
#' @export
which_r <- function(x, y) {
  n <- length(y)
  res <- c()
  j <- 0

  for (i in 1:n) {
    if (x(y[i])) {
      j <- j + 1
      res[j] <- i
    }
  }

  if (j == 0) {
    return(0)
  } else {
    return(res)
  }
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the index of the first TRUE element in a vector (C++)
#' @inheritParams which_r
#' @export
which_cpp <- function(x, y) {
  which_cpp_(x, y)
}
```

A benchmark of the two functions is the following.

```{r 04-functions-bench3}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

head(which(x > 2))
head(which_cpp(function(x) x > 2, x))
head(which_r(function(x) x > 2, x))

mark(
  which(x > 2),
  which_cpp(function(x) x > 2, x),
  which_r(function(x) x > 2, x)
)
```

Clearly, there is no advantage over `which()` in this case.

## References

* [Get started with cpp11](https://cran.r-project.org/web/packages/cpp11/vignettes/cpp11.html)
