# Logical Functions

```{r 04-setup}
#| label: setup
#| echo: false
#| message: false
load_all()
```

## Are Some Values True? (`any()`)

`any()` returns `TRUE` if there is any `TRUE` element in a vector, and `FALSE`
otherwise. Here is my C++ implementation.

```cpp
[[cpp11::register]] bool any_cpp_(logicals x) {
  int n = x.size();

  for(int i = 0; i < n; ++i) {
    if (x[i]) {
      return true;
    }
  }
  return false;
}
```

The R equivalent is the following.

```r
#' Return TRUE if any element in a vector is TRUE (R)
#' @param x logical vector
#' @export
any_r <- function(x) {
  n <- length(x)

  for (i in 1:n) {
    if (x[i]) {
      return(TRUE)
    }
  }
  FALSE
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return TRUE if any element in a vector is TRUE (C++)
#' @inheritParams any_r
#' @export
any_cpp <- function(x) {
  any_cpp_(x)
}
```

Benchmark.

```{r 04-functions-bench2}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements
y <- ifelse(x > 2, TRUE, FALSE)

any(y)
any_cpp(y)
any_r(y)

mark(
  any(y),
  any_cpp(y),
  any_r(y)
)
```

## Which indices are TRUE? (`which()`)

`which()` returns the indexes of the `TRUE` elements in a vector. Here is
my C++ implementation.

```cpp
[[cpp11::register]] integers which_cpp_(function pred, doubles x) {
  int n = x.size();
  writable::doubles res;
  int j = 0;

  for(int i = 0; i < n; ++i) {
    if (pred(x[i])) {
      ++j;
      res.push_back(i + 1);
    }
  }

  if (j == 0) {
    return doubles(0);
  } else {
    return res;
  }
}
```

The R equivalent is the following.

```r
#' Return the index of the first TRUE element in a vector (R)
#' @param x predicate function
#' @param y vector of values
#' @export
which_r <- function(x, y) {
  n <- length(y)
  res <- c()
  j <- 0

  for (i in 1:n) {
    if (x(y[i])) {
      j <- j + 1
      res[j] <- i
    }
  }

  if (j == 0) {
    return(0)
  } else {
    return(res)
  }
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the index of the first TRUE element in a vector (C++)
#' @inheritParams which_r
#' @export
which_cpp <- function(x, y) {
  which_cpp_(x, y)
}
```

A benchmark of the two functions is the following.

```{r 04-functions-bench3}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

head(which(x > 2))
head(which_cpp(function(x) x > 2, x))
head(which_r(function(x) x > 2, x))

mark(
  which(x > 2),
  which_cpp(function(x) x > 2, x),
  which_r(function(x) x > 2, x)
)
```

There is no advantage over `which()` in this case.

## Are all values true? (`all()`)

In R we can use `all()` to check if all values in a vector are `TRUE`, as in:

```{r 05-functions-all}
all(1:5 > 1)
``` 

Perhaps the easiest C++ implementation is to define a function that takes a
vector of logicals and returns a single logical value after the boolean result
of a loop over the length of the vector.

```cpp
[[cpp11::register]] bool all_cpp_1_(logicals x) {
  int n = x.size();
  for(int i = 0; i < n; ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
``` 

I can save lines by not defining `n`.

```cpp
[[cpp11::register]] bool all_cpp_2_(logicals x) {
  for (int i = 0; i < x.size(); ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
```

It is also possible to directly define a logical variable `i` inside the loop.

```cpp
[[cpp11::register]] bool all_cpp_3_(logicals x) {
  for (bool i : x) {
    if (i == false) {
      return false;
    }
  }
  return true;
}
```

Using `std::all_of()` I can do the same thing in one line.

```cpp
[[cpp11::register]] bool all_cpp_4_(logicals x) {
  return std::all_of(x.begin(), x.end(), [](bool x) { return x; });
}
```

To avoid typing `std::` I can use `using namespace std;` at the top of
`src/code.cpp`.

These functions can be directly called from R, in the previous part the
goal was to write a basic documentation and default parameters by using
auxiliary functions.

Now I need to test if the functions work as expected.

```{r 05-functions-all-correctness}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

all(x > 2)
all_cpp_1_(x > 2)
all_cpp_2_(x > 2)
all_cpp_3_(x > 2)
all_cpp_4_(x > 2)

# also test the TRUE only case
all(x >= 0)
all_cpp_1_(x >= 0)
all_cpp_2_(x >= 0)
all_cpp_3_(x >= 0)
all_cpp_4_(x >= 0)
```

Now I can care about the benchmarking.

```{r 05-functions-all-benchmark}
mark(
  all(x > 2),
  all_cpp_1_(x > 2),
  all_cpp_2_(x > 2),
  all_cpp_3_(x > 2),
  all_cpp_4_(x > 2)
)
```

My functions are marginally better (in terms of speed) than base R.

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
