---
format: gfm
editor:
  markdown:
    wrap: 80
---

# Functions (part 4)

## Are all values true?

In R we can use `all()` to check if all values in a vector are `TRUE`, as in:

```{r 05-functions-all}
all(1:5 > 1)
``` 

Perhaps the easiest C++ implementation is to define a function that takes a
vector of logicals and returns a single logical value after the boolean result
of a loop over the length of the vector.

```cpp
[[cpp11::register]] bool all_cpp_1_(cpp11::logicals x) {
  int n = x.size();
  for(int i = 0; i < n; ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
``` 

I can save lines by not defining `n`.

```cpp
[[cpp11::register]] bool all_cpp_2_(cpp11::logicals x) {
  for (int i = 0; i < x.size(); ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
```

It is also possible to directly define a logical variable `i` inside the loop.

```cpp
[[cpp11::register]] bool all_cpp_3_(cpp11::logicals x) {
  for (bool i : x) {
    if (i == false) {
      return false;
    }
  }
  return true;
}
```

Using `std::all_of()` I can do the same thing in one line.

```cpp
[[cpp11::register]] bool all_cpp_4_(cpp11::logicals x) {
  return std::all_of(x.begin(), x.end(), [](bool x) { return x; });
}
```

To avoid typing `std::` I can use `using namespace std;` at the top of
`src/code.cpp`.

These functions can be directly called from R, in the previous part the
goal was to write a basic documentation and default parameters by using
auxiliary functions.

Now I need to test if the functions work as expected.

```{r 05-functions-all-correctness}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

all(x > 2)
all_cpp_1_(x > 2)
all_cpp_2_(x > 2)
all_cpp_3_(x > 2)
all_cpp_4_(x > 2)

# also test the TRUE only case
all(x >= 0)
all_cpp_1_(x >= 0)
all_cpp_2_(x >= 0)
all_cpp_3_(x >= 0)
all_cpp_4_(x >= 0)
```

Now I can care about the benchmarking.

```{r 05-functions-all-benchmark}
mark(
    all(x > 2),
    all_cpp_1_(x > 2),
    all_cpp_2_(x > 2),
    all_cpp_3_(x > 2),
    all_cpp_4_(x > 2)
)
```

My functions are marginally better (in terms of speed) than base R.

## Cumulative product

In R we can use `cumprod()` to compute the cumulative product of a vector.

```{r 05-functions-cumprod}
cumprod(1:5)
```

The C++ implementation can use the shortcuts from the previous part.

```cpp
[[cpp11::register]] cpp11::doubles cumprod_cpp_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out(n);
  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] * x[i];
  }
  return out;
}
```

I can test the correctness of the function.

```{r 05-functions-cumprod-correctness, error = T}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

cumprod(1:5)
cumprod_cpp_(1:5)
```

I need an auxiliary function to cast the input as double.

```r
#' @export
cumprod_cpp <- function(x) {
  cumprod_cpp_(as.double(x))
}
```

Now I can test the correctness of the function again.

```{r 05-functions-cumprod-correctness2}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

cumprod(1:5)
cumprod_cpp(1:5)
```

Now I can benchmark the functions.

```{r 05-functions-cumprod-benchmark}
set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

mark(
    cumprod(x),
    cumprod_cpp(x)
)
```

The C++ implementation is slower than base R.

One remedial solution is not to define the length of the vector and add one coordinate on each step.

```cpp
[[cpp11::register]] cpp11::doubles cumprod_cpp_2_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out;
  out.push_back(x[0]);
  for(int i = 1; i < n; ++i) {
    out.push_back(out[i - 1] * x[i]);
  }
  return out;
}
```

The benchmark is.

```{r 05-functions-cumprod-benchmark2}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

library(ggplot2)
library(dplyr)
library(tidyr)

results <- mark(
    cumprod(x),
    cumprod_cpp(x),
    cumprod_cpp_(as.double(x)),
    cumprod_cpp_2_(as.double(x))
)

results %>%
    unnest(c(time, mem_alloc, gc)) %>%
    select(expression, time, mem_alloc, gc) %>%
    filter(gc == "none") %>%
    ggplot(aes(x = mem_alloc, y = time, color = expression)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(method = "lm", se = F, colour = "grey50") +
    theme_minimal()
```

`cumprod_cpp_2_` is faster than `cumprod_cpp_` but still slower than base R, and
it uses more memory than any of the compared function calls.

## Cumulative minimum

Let's look at this example from R's documentation.

```{r 05-functions-cummin}
c(3:1, 2:0, 4:2)

cummin(c(3:1, 2:0, 4:2))
```

The function starts with the first element and then it compares the next
element with the current minimum and keeps the smallest value. In this case,
when it reaches zero, the next values in the sequence are zeroes because there
are no negative values in the original vector.

The lesson from the previous part is to avoid growing vectors, it uses more
memory.

The C++ implementation can use the learning from the previous parts.

```cpp
[[cpp11::register]] cpp11::doubles cummin_cpp_(cpp11::doubles x) {
  int n = x.size();

  // create a vector of 0s
  writable::doubles out(n);

  for (int i = 0; i < n; ++i) {
    out[i] = 0.0;
  }

  out[0] += x[0];

  for (int i = 1; i < n; ++i) {
    if (x[i] < out[i - 1]) {
      out[i] += x[i];
    } else {
      out[i] += out[i - 1];
    }
  }
  return out;
}
```

I can test the correctness of the function.

```{r 05-functions-cummin-correctness}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

cummin(c(3:1, 2:0, 4:2))
cummin_cpp_(as.double(c(3:1, 2:0, 4:2)))
```

The benchmark is.

```{r 05-functions-cummin-benchmark}
# create random vectors with 1, 3, 5, 9 million elements
set.seed(123) # set the seed for reproducibility
x1 <- as.double(rpois(1e6, lambda = 2))
x2 <- as.double(rpois(3e6, lambda = 2))
x3 <- as.double(rpois(5e6, lambda = 2))
x4 <- as.double(rpois(7e6, lambda = 2))
x5 <- as.double(rpois(9e6, lambda = 2))

results <- purrr::map(
    list(x1, x2, x3, x4, x5),
    ~ mark(
        cummin(.x),
        cummin_cpp_(.x)
    ) %>%
        mutate(n = length(.x))
)

results %>%
    bind_rows() %>%
    unnest(c(time, mem_alloc, gc, n)) %>%
    select(expression, time, mem_alloc, gc, n) %>%
    filter(gc == "none") %>%
    mutate(expr = paste(expression, n, sep = " N: ")) %>%
    ggplot(aes(x = mem_alloc, y = time, color = expr)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(aes(color = expression), method = "lm", se = F) +
    theme_minimal()
```


## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)

