---
format: gfm
editor:
  markdown:
    wrap: 80
---

# Functions (part 4)

## Are all values true?

In R we can use `all()` to check if all values in a vector are `TRUE`, as in:

```{r 05-functions-all}
all(1:5 > 1)
``` 

Perhaps the easiest C++ implementation is to define a function that takes a
vector of logicals and returns a single logical value after the boolean result
of a loop over the length of the vector.

```cpp
[[cpp11::register]] bool all_cpp_1_(cpp11::logicals x) {
  int n = x.size();
  for(int i = 0; i < n; ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
``` 

I can save lines by not defining `n`.

```cpp
[[cpp11::register]] bool all_cpp_2_(cpp11::logicals x) {
  for (int i = 0; i < x.size(); ++i) {
    if (x[i] == false) {
      return false;
    }
  }
  return true;
}
```

It is also possible to directly define a logical variable `i` inside the loop.

```cpp
[[cpp11::register]] bool all_cpp_3_(cpp11::logicals x) {
  for (bool i : x) {
    if (i == false) {
      return false;
    }
  }
  return true;
}
```

Using `std::all_of()` I can do the same thing in one line.

```cpp
[[cpp11::register]] bool all_cpp_4_(cpp11::logicals x) {
  return std::all_of(x.begin(), x.end(), [](bool x) { return x; });
}
```

To avoid typing `std::` I can use `using namespace std;` at the top of
`src/code.cpp`.

These functions can be directly called from R, in the previous part the
goal was to write a basic documentation and default parameters by using
auxiliary functions.

Now I need to test if the functions work as expected.

```{r 05-functions-all-correctness}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

all(x > 2)
all_cpp_1_(x > 2)
all_cpp_2_(x > 2)
all_cpp_3_(x > 2)
all_cpp_4_(x > 2)

# also test the TRUE only case
all(x >= 0)
all_cpp_1_(x >= 0)
all_cpp_2_(x >= 0)
all_cpp_3_(x >= 0)
all_cpp_4_(x >= 0)
```

Now I can care about the benchmarking.

```{r 05-functions-all-benchmark}
mark(
  all(x > 2),
  all_cpp_1_(x > 2),
  all_cpp_2_(x > 2),
  all_cpp_3_(x > 2),
  all_cpp_4_(x > 2)
)
```

My functions are marginally better (in terms of speed) than base R.

## Cumulative product

In R we can use `cumprod()` to compute the cumulative product of a vector.

```{r 05-functions-cumprod}
cumprod(1:5)
```

The C++ implementation can use the shortcuts from the previous part.

```cpp
[[cpp11::register]] cpp11::doubles cumprod_cpp_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out(n);
  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] * x[i];
  }
  return out;
}
```

I can test the correctness of the function.

```{r 05-functions-cumprod-correctness, error = T}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

cumprod(1:5)
cumprod_cpp_(1:5)
```

I need an auxiliary function to cast the input as double.

```r
#' @export
cumprod_cpp <- function(x) {
  cumprod_cpp_(as.double(x))
}
```

Now I can test the correctness of the function again.

```{r 05-functions-cumprod-correctness2}
# just to make sure I am in the right folder when rendering the qmd file
setwd("~/github/cpp-for-r-users/ece244")

load_all()

cumprod(1:5)
cumprod_cpp(1:5)
```

Now I can benchmark the functions.

```{r 05-functions-cumprod-benchmark}
set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

mark(
  cumprod(x),
  cumprod_cpp(x)
)
```

The C++ implementation is slower than base R.

## References

* [Get started with cpp11](https://cran.r-project.org/web/packages/cpp11/vignettes/cpp11.html)
