# Rolling functions

## Cumulative sum (`cumsum()`)

To be able to create writable vectors, I need to append this below the namespace
declaration (i.e., `using namespace cpp11`).

```cpp
namespace writable = cpp11::writable;
```

This function returns the cumulative sum of the elements of a vector.

```cpp
[[cpp11::register]] doubles cumsum_cpp_(doubles x) {
  int n = x.size();
  writable::doubles out(n);

  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] + x[i];
  }
  return out;
}
```

The R equivalent is the following.

```r
#' Return the cumulative sum of the coordinates of a vector (R)
#' @param x numeric vector
#' @export
cumsum_r <- function(x) {
  n <- length(x)
  out <- numeric(n)
  out[1] <- x[1]
  for (i in 2:n) {
    out[i] <- out[i - 1] + x[i]
  }
  out
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the cumulative sum of the coordinates of a vector (C++)
#' @inheritParams cumsum_r
#' @export
cumsum_cpp <- function(x) {
  cumsum_cpp_(as.double(x))
}
```

A benchmark of the two functions is the following.

```{r 05-cumsum-bench}
load_all()

x <- runif(1e3) # 1,000,000 elements

cumsum(1:3)
cumsum_cpp(1:3)
cumsum_r(1:3)

mark(
    cumsum(x),
    cumsum_cpp(x),
    cumsum_r(x)
)
```

My C++ function is in the middle between R's `cumsum()` and my R function
`cumsum_r()`.

## Cumulative product (`cumprod()`)

In R we can use `cumprod()` to compute the cumulative product of a vector.

```{r 05-functions-cumprod}
cumprod(1:5)
```

The C++ implementation can use the shortcuts from the previous part.

```cpp
[[cpp11::register]] cpp11::doubles cumprod_cpp_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out(n);
  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] * x[i];
  }
  return out;
}
```

Test correctness.

```{r 05-functions-cumprod-correctness, error = T}
load_all()

cumprod(1:5)
cumprod_cpp_(1:5)
```

I need an auxiliary function to cast the input as double.

```r
#' Return the cumulative product of the coordinates of a vector (C++)
#' @param x numeric vector
#' @export
cumprod_cpp <- function(x) {
  cumprod_cpp_(as.double(x))
}
```

Now I can test the correctness of the function again.

```{r 05-functions-cumprod-correctness2}
load_all()

cumprod(1:5)
cumprod_cpp(1:5)
```

Benchmark.

```{r 05-functions-cumprod-benchmark}
set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

mark(
    cumprod(x),
    cumprod_cpp(x)
)
```

The C++ implementation is slower than base R.

One remedial solution is not to define the length of the vector and add one
coordinate on each step.

```cpp
[[cpp11::register]] cpp11::doubles cumprod_cpp_2_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out;
  out.push_back(x[0]);
  for(int i = 1; i < n; ++i) {
    out.push_back(out[i - 1] * x[i]);
  }
  return out;
}
```

Benchmark.

```{r 05-functions-cumprod-benchmark2}
load_all()

set.seed(123) # set the seed for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

library(ggplot2)
library(dplyr)
library(tidyr)

results <- mark(
    cumprod(x),
    cumprod_cpp(x),
    cumprod_cpp_(as.double(x)),
    cumprod_cpp_2_(as.double(x))
)

results %>%
    unnest(c(time, mem_alloc, gc)) %>%
    select(expression, time, mem_alloc, gc) %>%
    filter(gc == "none") %>%
    ggplot(aes(x = mem_alloc, y = time, color = expression)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(method = "lm", se = F, colour = "grey50") +
    theme_minimal()
```

`cumprod_cpp_2_` is faster than `cumprod_cpp_` but still slower than base R, and
it uses more memory than any of the compared function calls.

## Cumulative minimum (`cummin()`)

Let's look at this example from R's documentation.

```{r 05-functions-cummin}
c(3:1, 2:0, 4:2)

cummin(c(3:1, 2:0, 4:2))
```

The function starts with the first element and then it compares the next
element with the current minimum and keeps the smallest value. In this case,
when it reaches zero, the next values in the sequence are zeroes because there
are no negative values in the original vector.

The lesson from the previous part is to avoid growing vectors, it uses more
memory.

The C++ implementation can use the learning from the previous parts.

```cpp
[[cpp11::register]] cpp11::doubles cummin_cpp_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out(n);

  out[0] = x[0];
  for (int i = 1; i < n; ++i) {
    // error: taking address of rvalue [-fpermissive]
    // double* x1 = &out[i - 1];
    double x1 = out[i - 1];

    // error: lvalue required as unary ‘&’ operand
    // double* x2 = &x[i];
    double x2 = x[i];

    out[i] = std::min(x1, x2);
  }
  return out;
}
```

Test correctness and benchmark.

```{r 05-functions-cummin-correctness}
load_all()

cummin(c(3:1, 2:0, 4:2))
cummin_cpp_(as.double(c(3:1, 2:0, 4:2)))

# create random vectors
set.seed(123) # set the seed for reproducibility
x1 <- as.double(rpois(1e6, lambda = 2))
x2 <- as.double(rpois(10e6, lambda = 2))
x3 <- as.double(rpois(20e6, lambda = 2))

results <- purrr::map(
    list(x1, x2, x3),
    ~ mark(
        cummin(.x),
        cummin_cpp_(.x)
    ) %>%
        mutate(n = length(.x))
)

results %>%
    bind_rows() %>%
    unnest(c(time, mem_alloc, gc, n)) %>%
    select(expression, time, mem_alloc, gc, n) %>%
    filter(gc == "none") %>%
    mutate(expr = paste(expression, n, sep = " N: ")) %>%
    ggplot(aes(x = mem_alloc, y = time, color = expr)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(aes(color = expression), method = "lm", se = F) +
    theme_minimal()
```

## Cumulative maximum (`cummax()`)

Analogous to the previous part.

```cpp
[[cpp11::register]] cpp11::doubles cummax_cpp_(cpp11::doubles x) {
  int n = x.size();
  writable::doubles out(n);

  out[0] = x[0];
  for (int i = 1; i < n; ++i) {
    // error: taking address of rvalue [-fpermissive]
    // double* x1 = &out[i - 1];
    double x1 = out[i - 1];

    // error: lvalue required as unary ‘&’ operand
    // double* x2 = &x[i];
    double x2 = x[i];

    out[i] = std::max(x1, x2);
  }
  return out;
}
```

Test correctness and benchmark.

```{r 05-functions-cummax-correctness}
load_all()

cummax(c(3:1, 2:0, 4:2))
cummax_cpp_(as.double(c(3:1, 2:0, 4:2)))

# create random vectors
set.seed(123) # set the seed for reproducibility
x1 <- as.double(rpois(1e6, lambda = 2))
x2 <- as.double(rpois(10e6, lambda = 2))
x3 <- as.double(rpois(20e6, lambda = 2))

results <- purrr::map(
    list(x1, x2, x3),
    ~ mark(
        cummax(.x),
        cummax_cpp_(.x)
    ) %>%
        mutate(n = length(.x))
)

results %>%
    bind_rows() %>%
    unnest(c(time, mem_alloc, gc, n)) %>%
    select(expression, time, mem_alloc, gc, n) %>%
    filter(gc == "none") %>%
    mutate(expr = paste(expression, n, sep = " N: ")) %>%
    ggplot(aes(x = mem_alloc, y = time, color = expr)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(aes(color = expression), method = "lm", se = F) +
    theme_minimal()
```

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
