# Statistical functions with missing values

```{r 06-setup}
#| label: setup
#| echo: false
#| message: false
load_all()
```

## Sum of Vector Elements (`sum()`)

The following function returns the sum of the elements of a vector.

```cpp
[[cpp11::register]] double sum2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();
  double total = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      total += x[i];
    }
  }
  return total;
}
```

I also have to add the corresponding auxiliary function for the documentation.

```r
#' Return the sum of the coordinates of a vector (C++)
#' @inheritParams sum_r
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
sum2_cpp <- function(x, na_rm = FALSE) {
  sum2_cpp_(as.double(x), na_rm = na_rm)
}
```

To test, I run the following lines in the R console.

```{r 06-sum-test}
set.seed(123) # for reproducibility
x <- runif(1e3) # 1,000,000 elements
x[sample(1:1e3, 1e2)] <- NA # change some elements to NA at random

sum(x, na.rm = FALSE)
sum2_cpp(x, na_rm = FALSE)

sum(x, na.rm = TRUE)
sum2_cpp(x, na_rm = TRUE)

mark(
  sum(x, na.rm = TRUE),
  sum2_cpp(x, na_rm = TRUE)
)
```

## Arithmetic Mean (`mean()`)

This function returns the average (or mean) of the elements of a vector.

```cpp
[[cpp11::register]] double mean2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  int m = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      ++m;
    }
  }

  if (m == 0) {
    return NA_REAL;
  }

  double y = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      y += x[i];
    }
  }

  return y / m;
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the mean of the coordinates of a vector (C++)
#' @inheritParams sum2_cpp
#' @export
mean2_cpp <- function(x, na_rm = FALSE) {
  mean2_cpp_(as.double(x), na_rm = na_rm)
}
```

A benchmark of the two functions is the following.

```{r 06-mean-bench}
mean(x)
mean2_cpp(x)

mean(x, na.rm = TRUE)
mean2_cpp(x, na_rm = TRUE)

mark(
  mean(x, na.rm = TRUE),
  mean2_cpp(x, na_rm = TRUE)
)
```

## Variance (`var()`)

This function returns the variance of the elements of a vector.

```cpp
[[cpp11::register]] double var2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  if (n < 2) {
    return NA_REAL;
  }

  int m = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      ++m;
    }
  }

  if (m < 2) {
    return NA_REAL;
  }

  double ex = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      ex += x[i];
    }
  }
  ex /= m;

  double out = 0;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      out += pow(x[i] - ex, 2);
    }
  }

  return out / (m - 1);
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the variance of the coordinates of a vector (C++)
#' @inheritParams sum2_cpp
#' @export
var2_cpp <- function(x, na_rm = FALSE) {
  var2_cpp_(as.double(x), na_rm = na_rm)
}
```

A benchmark of the two functions is the following.

```{r 03-var-bench}
var(x)
var2_cpp(x)

var(x, na.rm = TRUE)
var2_cpp(x, na_rm = TRUE)

mark(
  var(x, na.rm = TRUE),
  var2_cpp(x, na_rm = TRUE)
)
```

## Root Mean Square Error

The next function returns the measure of the differences between the observed
values or an estimator (x1, x2, ..., xn) and the true value (x0). For example,
x1, ..., xn could be experimental averages and x0 the true average.

```cpp
[[cpp11::register]] double rmse2_cpp_(doubles x, double x0) {
  int n = x.size();

  int m = 0;
  for (int i = 0; i < n; ++i) {
    if (ISNAN(x[i])) {
      continue;
    } else {
      ++m;
    }
  }

  if (m == 0) {
    return NA_REAL;
  }

  double out;
  for (int i = 0; i < n; ++i) {
    if (na_rm && ISNAN(x[i])) {
      continue;
    } else {
      out += pow(x[i] - x0, 2.0);
    }
  }
  return sqrt(out / m);
}
```

I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the root mean square error (C++)
#' @inheritParams rmse_r
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
rmse2_cpp <- function(x, x0, na_rm = FALSE) {
  rmse2_cpp_(as.double(x), as.double(x0), na_rm = na_rm)
}
```

A benchmark of the base R versus C++ implementation is the following.

```{r 03-functions-bench}
# create a list with 100 normal distributions with mean 0 and 1 million elements
# each
set.seed(123) # for reproducibility
x <- list()
for (i in 1:1e3) {
  x[[i]] <- rnorm(1e3)
}

# compute the mean of each distribution
x <- sapply(x, mean)

# remove some elements at random
x[sample(1:1e3, 1e2)] <- NA

rmse2_cpp(x, 0)
rmse2_cpp(x, 0, na_rm = TRUE)

mark(
  sqrt(mean((x - 0)^2, na.rm = TRUE)),
  rmse2_cpp(x, 0, na_rm = TRUE)
)
```

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
