# Rolling functions with missing values

```{r 05-setup}
#| label: setup
#| echo: false
#| message: false
load_all()
```

## Required packages

Needed to visualize and simplify showing the results.

```{r 05-packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
```

## Cumulative sum (`cumsum()`)

This function returns the cumulative sum of the elements of a vector.

```cpp
[[cpp11::register]] doubles cumsum2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  if (na_rm == true) {
    writable::doubles x2;
    for (int i = 0; i < n; ++i) {
      if (ISNAN(x[i])) {
        continue;
      } else {
        x2.push_back(x[i]);
      }
    }
    
    int m = x2.size();
    writable::doubles out;
    if (m == 0) {
      out.push_back(NA_REAL);
      return out;
    }

    out.push_back(x2[0]);
    for (int i = 1; i < m; ++i) {
      out.push_back(out[i - 1] + x2[i]);
    }

    return out;
  } else {
    writable::doubles out(n);
    if (ISNAN(x[0])) {
      out[0] = NA_REAL;
    } else {
      out[0] = x[0];
    }
    for (int i = 1; i < n; ++i) {
      if (ISNAN(x[i])) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(out[i - 1])) {
          out[i] = NA_REAL;
        } else {
          out[i] = out[i - 1] + x[i];
        }
      }
    }

    return out;
  }
}
```
I also need to add the corresponding auxiliary function for the documentation.

```r
#' Return the cumulative sum of the coordinates of a vector (C++)
#' @param x numeric vector
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
cumsum2_cpp <- function(x, na_rm = FALSE) {
  cumsum2_cpp_(as.double(x), na_rm = na_rm)
}
```

A benchmark of the two functions is the following.

```{r 05-cumsum-bench}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

cumsum(c(1, NA, 2, 4))
cumsum2_cpp(c(1, NA, 2, 4))
cumsum2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cumsum(x),
  cumsum2_cpp(x)
)
```

## Cumulative product (`cumprod()`)

The C++ implementation is similar to the previous part.

```cpp
[[cpp11::register]] doubles cumprod2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  if (na_rm == true) {
    writable::doubles x2;
    for (int i = 0; i < n; ++i) {
      if (ISNAN(x[i])) {
        continue;
      } else {
        x2.push_back(x[i]);
      }
    }

    int m = x2.size();
    writable::doubles out;
    if (m == 0) {
      out.push_back(NA_REAL);
      return out;
    }

    out.push_back(x2[0]);
    for (int i = 1; i < m; ++i) {
      out.push_back(out[i - 1] * x2[i]);
    }

    return out;
  } else {
    writable::doubles out(n);
    if (ISNAN(x[0])) {
      out[0] = NA_REAL;
    } else {
      out[0] = x[0];
    }
    for (int i = 1; i < n; ++i) {
      if (ISNAN(x[i])) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(out[i - 1])) {
          out[i] = NA_REAL;
        } else {
          out[i] = out[i - 1] * x[i];
        }
      }
    }

    return out;
  }
}
```

I need an auxiliary function to cast the input as double.

```r
#' Return the cumulative product of the coordinates of a vector (C++)
#' @inheritParams cumsum_r
#' @export
cumprod2_cpp <- function(x, na_rm = FALSE) {
  cumprod2_cpp_(as.double(x), na_rm = na_rm)
}
```

Test correctness and Benchmark.

```{r 05-functions-cumprod-benchmark}
cumprod(c(1, NA, 2, 4))
cumprod2_cpp(c(1, NA, 2, 4))
cumprod2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cumprod(x),
  cumprod2_cpp(x)
)
```

## Cumulative minimum (`cummin()`)

The C++ implementation is similar to the previous part.

```cpp
[[cpp11::register]] doubles cummin2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  if (na_rm == true) {
    writable::doubles x2;
    for (int i = 0; i < n; ++i) {
      if (ISNAN(x[i])) {
        continue;
      } else {
        x2.push_back(x[i]);
      }
    }

    int m = x2.size();
    writable::doubles out;
    if (m == 0) {
      out.push_back(NA_REAL);
      return out;
    }

    out.push_back(x2[0]);
    for (int i = 1; i < m; ++i) {
      double y1 = x2[i - 1], y2 = x2[i];
      out.push_back(std::min(y1, y2));
    }

    return out;
  } else {
    writable::doubles out(n);
    if (ISNAN(x[0])) {
      out[0] = NA_REAL;
    } else {
      out[0] = x[0];
    }
    for (int i = 1; i < n; ++i) {
      if (ISNAN(x[i])) {
        out[i] = NA_REAL;
      } else {
        double y1 = out[i - 1], y2 = x[i];
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = std::min(y1, y2);
        }
      }
    }

    return out;
  }
}
```

I also have to add the corresponding auxiliary function for the documentation.

```r
#' Return the cumulative minimum of the coordinates of a vector (C++)
#' @inheritParams cumsum_r
#' @export
cummin2_cpp <- function(x, na_rm = FALSE) {
  cummin2_cpp_(as.double(x), na_rm = na_rm)
}
```

Test correctness and benchmark.

```{r 05-cummin-correctness}
cummin(c(1, NA, 2, 4))
cummin2_cpp(c(1, NA, 2, 4))
cummin2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cummin(x),
  cummin2_cpp(x)
)
```

## Cumulative maximum (`cummax()`)

The C++ implementation is similar to the previous part.

```cpp
[[cpp11::register]] doubles cummin2_cpp_(doubles x, bool na_rm = false) {
  int n = x.size();

  if (na_rm == true) {
    writable::doubles x2;
    for (int i = 0; i < n; ++i) {
      if (ISNAN(x[i])) {
        continue;
      } else {
        x2.push_back(x[i]);
      }
    }

    int m = x2.size();
    writable::doubles out;
    if (m == 0) {
      out.push_back(NA_REAL);
      return out;
    }

    out.push_back(x2[0]);
    for (int i = 1; i < m; ++i) {
      double y1 = x2[i - 1], y2 = x2[i];
      out.push_back(std::min(y1, y2));
    }

    return out;
  } else {
    writable::doubles out(n);
    if (ISNAN(x[0])) {
      out[0] = NA_REAL;
    } else {
      out[0] = x[0];
    }
    for (int i = 1; i < n; ++i) {
      if (ISNAN(x[i])) {
        out[i] = NA_REAL;
      } else {
        double y1 = out[i - 1], y2 = x[i];
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = std::min(y1, y2);
        }
      }
    }

    return out;
  }
}
```

I also have to add the corresponding auxiliary function for the documentation.

```r
#' Return the cumulative maximum of the coordinates of a vector (C++)
#' @inheritParams cumsum_r
#' @export
cummax2_cpp <- function(x, na_rm = FALSE) {
  cummax2_cpp_(as.double(x), na_rm = na_rm)
}
```

Test correctness and benchmark.

```{r 05-cummax-correctness}
cummax(c(1, NA, 2, 4))
cummax2_cpp(c(1, NA, 2, 4))
cummax2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cummax(x),
  cummax2_cpp(x)
)
```

## References

* [Get started with cpp11](https://cpp11.r-lib.org/articles/cpp11.html#intro)
